**Effective application of machine learning to a microbiome-based classification problem**

### 2020-01-20 submission/manuscript.Rmd
update for rebuttal

__Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. Because we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease, model interpretation becomes crucial for microbiome studies. ML models often decrease in interpretability as they increase in complexity. In this study, we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN. It was encouraging that many of the highest-ranked OTUs were shared across these three models (e.g., OTUs 50, 426, 609, 822, 1239). The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU. However, this approach is limited to linear models or models with prespecified interaction terms.

@@

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures that create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

**Figure S5. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

>>>>

__Interpretation of each ML model.__  To evaluate the overall structure of the CRC dataset and show that the structure of the human microbiome is remarkably variable among individuals with SRNs, we analyzed the Bray-Curtis distances with an NMDS ordination [Figure S5]. The lack of seperation between healthy patients and patients with SRNs, show that traditional statistical approaches do not help us to adequately address disease ecology. This is why we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease. Therefore, model interpretation becomes crucial for microbiome studies. Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. ML models often decrease in interpretability as they increase in complexity. In this study, we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN. It was encouraging that many of the highest-ranked OTUs were shared across these three models (e.g., OTUs 50, 426, 609, 822, 1239). The great benefit of this approach was knowing the sign and magnitude of each OTU coefficient in the trained model. This allowed us to immidiately interpret negative and positive coefficient signs as protective and risk factors, respectively and the magnitude as the impact of these factors. However, this approach is limited to linear models or models with prespecified interaction terms.

@@

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S6]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

We also compared the top 20 OTUs selected by permutation importance in L2-regularized logistic regression and the highest performing tree-based models, random forest and XGBoost. 2 and 5 OTUs, respectively, were shared among the models. These results indicate that we were able to identify a few more OTUs that are deemed important by both linear and non-linear models when we used permutation importance as our interpretation method.

Figure S7:

@@

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures that create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features. We are choosing to leave the decision to define correlation and setting up correlations thresholds to the researchers for their own microbiome questions. However, we would like to urge them to investigate the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. True causal mechanisms (e.g. role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020]. To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

**Figure S5. NMDS ordination with Bray-Curtis distances.**

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}

**Figure S6. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
